Higher-Order unification for free!

### 2024-04-16 paper/main.tex
update tex

The simple compiler described in this section serves as a base for the
extensions in \cref{sec:llam,sec:eta,sec:beta}.
Its main task is to beta normalize the term and map one syntax tree to
the other.
In order to bring back the substitution from \Ho{} to \Fo{} the compiler
builds a ``memory map'' connecting the kind of variables using routine
\ref{clause:malloc}.\noteD{non mi piace}

@@

\noindent
The code above uses that possibility
in order to allocate space for the variables, i.e. it sets their memory
address to \elpiIn{none} (a details not worth mentioning in the
previous sections).
\todo{explain fold6}

@@

% \noteD[inline]{Below the proof of \cref{prop:w-preservation}, ho usato 3 lemmi ausiliari,
% forse si può compattare in una prova più piccola?}
\begin{lemma}
  Given a \linketa $l$, the unification done by \progressetaleft is between
  terms in $\wellb$  
  \label{lemma:unif-eta-1}
\end{lemma}

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
  \todo{is w sigmaa defined?}
  \cref{prop:w-preservation}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
  \progressetaleft, if 1) the \lhs is a name, a constant or an
  application, then, $\lambda x.\lhs \appsep x$ is unified with the \rhs. By
  \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux},
  $\rhs = \lambda x. t$ and $\wellb(t)$.
  Otherwise, 2) the \lhs has \elpiIn{lam} as functor.
  \todo{what does as a functor mean?}
  In both cases, unification is
  performed between terms in \wellb.
\end{proof}

@@

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, if \noteD{riprendere sec 3.1}
  $\wellb_{\beta \eta}(\foUnifPb)$ and if the 
  memory map is bijective, then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{This premise is fulfilled in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}
\todo{should we introduce W/eta as prev section W/beta?}

\begin{proof}[Proof sketch]
  \progressetaleft and \progressetadedup activate a \linketa when, in the
  original unification problem, a \maybeeta term is unified with respectively a
  well-behaved term or another \maybeeta term. In both cases, the links trigger
  a unification which succeeds iff the same unification in \Fo{} succeeds,
  \todo{Fix: show that 5.7 an 5.8 compensate for missing eta rule}
  guaranteeing \cref{prop:fidelity}. \progressetaright never fails, in fact,
  this progression refines a variable to a rigid term and plays no role in 
  \cref{prop:fidelity}.
  % If 1) a \linketa is activated by \progressetaleft, then unification is done
  % between a \maybeeta term $t_1$ and a term $t_2$ (with $\wellb(t_2)$). This
  % activation performs a unification which succeeds iff the original problem in
  % \Fo{} succeeds. If 2) a \linketa is activated by \progressetadedup, then the
  % unification is done between two \maybeeta terms, and again this unification
  % succeeds iff it succeeds in \Fo{}. Finally, if 3) a \linketa is activated by
  % \progressetaright, the unification, done between a variable and a term,
  % always succeeds, this is what we expect to guarantee fidelity, since we
  % are essentially removing a hole added by the compilation in the place of a
  % \maybeeta subterm. In all the cases fidelity is respected.
\end{proof}

@@

\begin{theorem}[Fidelity with \mapdeduplication]
  Given a list of unification problems \foUnifPb, such that 
  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow \wellb(t) \lor t \in \maybeeta$,
  if $\;\foUnifPb$ contains the same \Fo{} variable used at different arities,
  then \mapdeduplication guarantees \fullref{prop:fidelity}
\end{theorem}

\begin{proof}[Proof sketch]
  By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
  variables $X_1, X_2$ with different arities are compiled to two
  variables $A^n$ and $C^m$ related by a chain of \linketa.
  We show that whenever one of the two is instantiated with a rigid term the 
  the chain becomes shorter and hence, eventually, the
  other variable is instantiated too.
  Without loss of generality we assume $m < n$ and $A$ is instantiated.
  The corresponding \linketa is triggered by \progressetaleft instantiating
  the next variable in the chain of links.
\end{proof}
\todo{Not sure I fullfy fixed the proof, but the old oen was not ok}

@@

\noindent
The resolution of $\hoUnifPb_1$ assigns $\lambda x.\lambda y.x$ to $A$
that in turn triggers $\linkStore_3$ and then $\linkStore_4$ by \progressetaleft,
one per $\lambda$ in the solution of $A$.
The unification variable $C_{yx}$ is therefore unified with $x$
(the second variable of its scope).
As a result $\linkStore_1$ becomes \linketaM{x}{E_x}{\lambda y.y},
and the \rhs is no more in \maybeeta so \progressetaright fires and
$E_x$ is unified with $\lambda y.y$, resulting in
$\sigma = \{~ A \mapsto \lambda x.\lambda y.x ~;~ B_x \mapsto \lambda y.x ~;~ C_{yx} \mapsto x ~;~ E_x \mapsto \lambda y.y~\}$.\\
The remaining step $\hoUnifPb_2$ identifies $D$ with $F$,
hence the resulting \Fo{} substitution is
\todo{verificare, la vecchia soluzione era sballata}
$\rho = \{X \mapsto \lambda x.\lambda y.y ~;~ Y \mapsto \lambda x.(f\appsep (\lambda y.y))\appsep x\}$.

@@

\noindent
The second snippet illustrate the deduplication of \linketa. The syntax
<<\elpiIn{N ~$\triangleright$~ G ?- P}>> denotes a $\lambda$Prolog sequent, that is a 
goal \elpiIn{P} under a set \elpiIn{G} of hypothetical rules (introduced
by \elpiIn{=>}) and where the program context binds (via the \elpiIn{pi}
operator) \elpi{N} heigenvariables. Sequents are presented to
CHR with their higher order unification variables replaced by
``frozen constants'', that is $A_{xyz}$ becomes <<\elpiIn{uvar ~$f_A$~ [x,y,z]}>>
for a fresh constant $f_A$. Frozen constants are ``defrost'' when
they are part of a new goal (see \cite[section 4.3]{TASSI_2019}).

>>>>

The simple compiler described in this section serves as a base for the
extensions in \cref{sec:llam,sec:eta,sec:beta}.
Its main task is to beta normalize the term and map one syntax tree to
the other.
In order to bring back the substitution from \Ho{} to \Fo{} the compiler
builds a ``memory map'' connecting the \Fo{} variables to the \Ho ones
using the routine
\ref{clause:malloc}.

@@

\noindent
The code above uses that possibility
in order to allocate space for the variables, i.e. it sets their memory
address to \elpiIn{none} (a detail not worth mentioning in the
previous sections).
\todo{explain fold6}

@@

% \noteD[inline]{Below the proof of \cref{prop:w-preservation}, ho usato 3 lemmi ausiliari,
% forse si può compattare in una prova più piccola?}
\begin{lemma}
  Given a \linketa \linketaM{\Gamma}{X}{T}, the unification done by \progressetaleft is between
  terms in $\wellb$  
  \label{lemma:unif-eta-1}
\end{lemma}

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, such that $\wellb(\sigma X)$ (since \Ue preserves \wellb),
  By\todo{reformulate}
  \progressetaleft, if 1) the $\sigma X$ is a name, a constant or an
  application, then, $\lambda x.X \appsep x$ is unified with the \rhs. By
  \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux},
  $\rhs = \lambda x. t$ and $\wellb(t)$.
  Otherwise, 2) $\sigma X = \lambda x.t$.
  % \todo{what does as a functor mean?}
  In both cases, unification is
  performed between terms in \wellb.
\end{proof}

@@

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, if
  $\wellb_{\beta \eta}(\foUnifPb)$ and if the 
  memory map is bijective, then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{This premise is fulfilled in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

\begin{proof}[Proof sketch]
  \progressetaleft and \progressetadedup activate a \linketa when, in the
  original unification problem, a \maybeeta term is unified with respectively a
  well-behaved term or another \maybeeta term. In both cases, the links trigger
  a unification which succeeds iff the same unification in \Fo{} succeeds,
  \todo{Fix: show that 5.7 an 5.8 compensate for missing eta rule, citare regolae in =o}
  guaranteeing \cref{prop:fidelity}. \progressetaright never fails, in fact,
  this progression refines a variable to a rigid term and plays no role in 
  \cref{prop:fidelity}.
  % If 1) a \linketa is activated by \progressetaleft, then unification is done
  % between a \maybeeta term $t_1$ and a term $t_2$ (with $\wellb(t_2)$). This
  % activation performs a unification which succeeds iff the original problem in
  % \Fo{} succeeds. If 2) a \linketa is activated by \progressetadedup, then the
  % unification is done between two \maybeeta terms, and again this unification
  % succeeds iff it succeeds in \Fo{}. Finally, if 3) a \linketa is activated by
  % \progressetaright, the unification, done between a variable and a term,
  % always succeeds, this is what we expect to guarantee fidelity, since we
  % are essentially removing a hole added by the compilation in the place of a
  % \maybeeta subterm. In all the cases fidelity is respected.
\end{proof}

@@

\begin{theorem}[Fidelity with \mapdeduplication]
  Given a list of unification problems \foUnifPb, such that 
  $\wellb_{\beta \eta}(\foUnifPb)$,
  if $\;\foUnifPb$ contains the same \Fo{} variable used at different arities,
  then \mapdeduplication guarantees \fullref{prop:fidelity}
\end{theorem}

\begin{proof}[Proof sketch]
  By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
  variables with different arities are compiled to two
  variables $A^n$ and $C^m$ related by a chain of \linketa.
  We show that whenever one of the two is instantiated with a rigid term then
  the chain becomes shorter and hence, eventually, the
  other variable is instantiated too.
  Without loss of generality we assume $m < n$ and $A$ is instantiated.
  The corresponding \linketa is triggered by \progressetaleft instantiating
  the next variable in the chain of links.
\end{proof}
\todo{Not sure I fullfy fixed the proof, but the old oen was not ok}

@@

\noindent
The resolution of $\hoUnifPb_1$ assigns $\lambda x.\lambda y.x$ to $A$
that in turn triggers $\linkStore_3$ and then $\linkStore_4$ by \progressetaleft,
one per $\lambda$ in the solution of $A$.
The unification variable $C_{yx}$ is therefore unified with $x$
(the second variable of its scope).
As a result $\linkStore_1$ becomes \linketaM{x}{E_x}{\lambda y.y},
and the \rhs is no more in \maybeeta so \progressetaright fires and
$E_x$ is unified with $\lambda y.y$; then $\linkStore_2$ makes progress thanks to the same 
rule giving
$\sigma = \{~ A \mapsto \lambda x.\lambda y.x ~;~ B_x \mapsto \lambda y.x ~;~ C_{yx} \mapsto x ~;~ D \mapsto f\appsep (\lambda y.y) ~;~ E_x \mapsto \lambda y.y ~\}$.\\
The remaining step $\hoUnifPb_2$ identifies $D$ with $F$,
hence the resulting \Fo{} substitution is
$\rho = \{X \mapsto \lambda x.\lambda y.y ~;~ Y \mapsto f\appsep (\lambda y.y)\}$.

@@

\noindent
The second snippet illustrate the deduplication of \linketa. The syntax
<<\elpiIn{N ~$\triangleright$~ G ?- P}>> denotes a $\lambda$Prolog sequent, that is a 
goal \elpiIn{P} under a set \elpiIn{G} of hypothetical rules (introduced
by \elpiIn{=>}) and where the program context binds (via the \elpiIn{pi}
operator) \elpiIn{N} heigenvariables. Sequents are presented to
CHR with their higher order unification variables replaced by
``frozen constants'', that is $A_{xyz}$ becomes <<\elpiIn{uvar ~$f_A$~ [x,y,z]}>>
for a fresh constant $f_A$. Frozen constants are ``defrost'' when
they are part of a new goal (see \cite[section 4.3]{TASSI_2019}).
